Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-07-19 Thesis/Conclusion.tex
Night before submission. Likely final pre-submission draft. Checked for spelling issues.

One of the key contributions of this thesis is the understanding of how the dynamic nature of stellar disks affects halo properties. Although the material in Chapter~\ref{ch:paper_i} is primarily focused on introducing a method used by our group, its key scientific contributions focus on this point.

We find that modelling the disk in a dynamic fashion is tremendously important for the inner halo. Specifically, we found the formation of a bar led to a more concentrated halo mass distribution. On the whole, these results suggest that if one is interested in the Core-Cusp problem, this should be studied with fully live models of stellar disks.

The intermediate-to-outer halo was less effected by the dynamics of the inner halo. While some differences between our disk potential models were observed in the halo mass functions, these differences were minor. The details of how satellites are disrupted appear to be at least weakly dependent on how the disk is modelled, and stream studies should consider as realistic of a disk as possible.

This last aspect of our work has informed some design decisions by others, namely in the work of \citet{read_2019}. If one is just interested in subhalo population statistics, it is sufficient to model baryons as fixed potentials in a pure dark matter halo.

Beyond the effect of the disk on the halo, we find that our rigid disk approximation is quite realistic in modelling the angular evolution of the disk. This allows us to set up near-equilibrium cosmological systems without the use of MHD simulations. Since the work in this thesis focuses on the underlying theoretical mechanisms driving the evolution of galaxies in cosmological environments, we only need a good representation of the true systems.

\subsection{Future Work}

Our algorithm makes some optimizations over previous disk insertion schemes. The optimizations we made are by no means exhaustive, and a detailed discussion of potential optimizations is given in Chapter~\ref{ch:implementation}. We should consider a scheme which extracts mass from the halo particles during the disk growth phase. Dark matter particles in a dark matter only simulation include the baryonic mass of the Universe. Since the disk and halo contribute on the same order to the potential in the inner parts of a galaxy, it would be more realistic to subtract mass from halo particles as the disk grows. This may have a sizable impact on our conclusions about the disk's effect on adiabatic contraction.

Finally, we might also consider a scheme which initializes the disk with some gas component. \citet{deg_2019} modified \textsc{GalactICS} to allow for the inclusion of gas. This, combined with modifying halo particle masses, should allow us to more accurately compare our results to fully hydrodynamical \textit{ab initio} simulations.

Even with the current algorithm, there is more we can do on the study of disk-halo interactions. A possible project would be to study the effect of varying disk properties in the same halo. We might also consider adding a spheroidal bulge component to see how high baryonic mass concentrations change our results. We might also look for more detailed evidence of the halo's response to the disk in our existing simulations.

\section{Bar Formation in Cosmological Stellar Disks}
\subsection{Summary}

We discuss our contributions to understanding how bars form in stellar disks in Chapter~\ref{ch:paper_ii}. The paper unifies previous work on disk insertion schemes by viewing them through a lens of a common parameter space. In doing so, we illustrate that various discrepancies between our work and previous work are actually due to selection effects in this space.

In particular, we show that a commonly discounted variable, the disk thickness, is a primary driver of bar strength in a cosmological setting. We show that, in general, the length of a bar at present day for cosmological stellar disks is largely a function of the initial disk thickness. This effect is not numerical, as we show through our application of the \textsc{AGAMA} code.

@@

A key flaw of the work in Chapter~\ref{ch:paper_ii} is that we only considered a single halo. In Chapter~\ref{ch:paper_iii}, we ran twelve simulations in four halos. Although all of the halos had the same total mass as the Milky Way, they varied widely in terms of their other properties. Despite the differences in the halos, the bars formed by embedded disks were quite similar. Additionally, there is a nontrivial dependence on Toomre $Q$ in terms of when the bar buckles, but the final bar strength itself does not depend on this parameter.

\subsection{Future Work}

In Chapter~\ref{ch:paper_iii}, bar formation was not the focus of the study. Now that we have a sample of bars forming in different cosmological environments, we can position the disks in the parameter space described in Chapter~\ref{ch:paper_ii}. A proper synthesis with previous literature on this topic should be performed.

An interesting feature of our simulations was the existence of ``banana bars" in some of the simulations. Understanding why we get banana bars in some simulations versus a more traditional X-shaped pattern might provide an observational constraint for the Milky Way. This topic should be explored in more detail.

A valid criticism of our disk insertion technique is that all of the stars are initialized simultaneously. This is unrealistic, and allowing stars to gradually form out of the rigid disk potential might show a more complicated dependence on the initial conditions parameter space. We think developments in this direction would be more directly comparable to \textit{ab initio} simulations, and present a vector for future research.

\section{Vetical Structure in Cosmological Stellar Disks}
\subsection{Summary}

The main focus of Chapter~\ref{ch:paper_iii} was how disks form vertical structure in $\Lambda$CDM environments. We showed that there are a wide variety of Milky Way-mass halos that form vertical structure consistent with structure observed in the solar neighborhood. In particular, we showed that this structure can be formed without a massive satellite encounter in a cosmological setting. This position is bolstered by the fact that we see Monoceros-like structure in all twelve of our simulations.

We also found that consistent perpendicular torques on the disk are required to explain features in the outer disk. It is unlikely that substructures far below the $10^{11}$ solar mass regime are responsible for stars stripped to higher latitudes. Global tides from disk-halo axis misalignment seem to be the most effective way to create these populations of stars. We therefore suggest that streams originating from the outer disk may primarily depend on the Milky Way's interaction with the LMC. A massive merger like the one with the LMC can cause consistent perpendicular torques its own tides or by disturbing the existing smooth Milky Way halo mass.

\subsection{Future Work}

A key flaw in this work was that we had no LMC analogues. An interesting experiment would be to compare the vertical structure caused by a future LMC merger and the vertical structure generated when a disk is misaligned in a triaxial halo. On the whole, future work should proceed in the same way as the work in Chapter~\ref{ch:paper_ii}. We should seek to relate the halo environments to a common parameter space.

@@

>>>>

One of the key contributions of this thesis is the understanding of how the dynamic nature of stellar disks affects halo properties, as studied in Chapter~\ref{ch:paper_i}. Some halo properties, such as the distribution of mass in the inner halo, depend a lot on how the disk is modelled. This is very evident in the differences in adiabatic contraction between the rigid and live disks. Specifically, we find the formation of a bar led to a more concentrated halo mass distribution. On the whole, these results suggest that if one is interested in the Core-Cusp problem, this should be studied with fully live models of stellar disks.

The intermediate-to-outer halo is less effected by the dynamics of the inner halo. While some differences between our disk potential models were observed in the halo mass functions, these differences were minor. The details of how satellites are disrupted appear to be at least weakly dependent on how the disk is modelled, and stream studies should consider as realistic of a disk as possible. This last aspect of our work has informed some design decisions by others, namely in the work of \citet{read_2019}. If one is just interested in subhalo population statistics, it is sufficient to model baryons as fixed potentials in a pure dark matter halo.

The global picture painted by our work is that modelling the disk as a rigid body does well in determining how the disk settles (i.e. its angular motion), a result consistent with \citet{DubinskiKuijkenRigidDisks}. It appears that the formation of a bar in the live disk is the most substantial difference between the rigid and live disks' effects on the inner halo, a difference that only becomes significant \textit{after} the the live disk

%Beyond the effect of the disk on the halo, we find that our rigid disk approximation is quite realistic in modelling the angular evolution of the disk. This allows us to set up near-equilibrium cosmological systems without the use of MHD simulations. Since the work in this thesis focuses on the underlying theoretical mechanisms driving the evolution of galaxies in cosmological environments, we only need a good representation of the true systems.

\subsection{Insertion Scheme and Future Work} \label{ssec:conclusions_insertion}

Our algorithm makes optimizations over previous disk insertion schemes. The optimizations we made are by no means exhaustive, and a detailed discussion of potential optimizations is given in Chapter~\ref{ch:implementation}. We focus here on scientific concerns with our algorithm, rather than performance optimizations.

The scheme as specified has the undesirable property of adding angular momentum and mass to the halo. This happens when the disk grows, and we fail to extract mass and angular momentum from the halo\footnote{Note that dark matter particles dark matter only simulations include the baryon mass of the Universe.}. A potential improvement which extracts mass and angular momentum from the halo particles during the disk growth phase would alleviate these concerns. This may have a sizable impact on our conclusions about the disk's effect on adiabatic contraction, since both the disk and halo are dominant in the inner galaxy.

We might also consider a scheme which initializes the disk with some gas component. \citet{deg_2019} modified \textsc{GalactICS} to allow for the inclusion of gas. This, combined with modifying halo particle masses, should allow us to more accurately compare our results to fully hydrodynamical \textit{ab initio} simulations.

A valid criticism of our disk insertion technique is that all of the stars are initialized simultaneously. This is unrealistic, and allowing stars to gradually form out of the rigid disk potential might show a more complicated dependence on the initial conditions parameter space. In fact, \citet{aumer_2016} grow disks in a similar fashion and find that slowly growing the disk along with gas physics impacts important properties like bar formation. In some cases it may even shut off bar formation, which we believe would substantially impact our conclusions about adiabatic contraction.

Even with the current algorithm, there is more we can do on the study of disk-halo interactions. A possible project would be to study the effect of varying disk properties in the same halo. We might also consider adding a spheroidal bulge component to see how high baryonic mass concentrations change our results. We might also look for more detailed evidence of the halo's response to the disk in our existing simulations in the form of wakes. This kind of response was explored in one of our toy models in Chapter~\ref{ch:paper_iii}, and it seems natural to extend this analysis to our cosmological simulations.

\section{Bar Formation}
\subsection{Summary}

We discuss our contributions to understanding how bars form in stellar disks in Chapter~\ref{ch:paper_ii}. The paper unifies previous work on disk insertion schemes by viewing them through a lens of a common parameter space. Through placing our models and the models of \citet{YurinSpringelStellarDisks} in this space, we are able to contextualize the very large bars formed in their study. In particular, we show that a commonly discounted variable, the disk thickness, is a primary driver of bar strength in a cosmological setting. We show that, in general, the length of a bar at present day for cosmological stellar disks is largely a function of the initial disk thickness. This effect is not numerical, as we show through our application of the \textsc{AGAMA} code \citep{vasiliev_2018}.

@@

We manage to bolster this result in Chapter~\ref{ch:paper_iii}, where we ran twelve simulations in four halos. Although all of the halos had the same total mass as the Milky Way, they varied widely in terms of their other properties. Despite the differences in the halos, the bars formed by embedded disks were quite similar. Additionally, there is a nontrivial dependence on Toomre $Q$ in terms of when the bar buckles, but the final bar strength depends very weakly on this parameter, contrary to \citet{YurinSpringelStellarDisks}. Rather, the trend \citet{YurinSpringelStellarDisks} sees in $Q$ is likely due to the presence of their bulge simulations, which would increase $Q$ as well as provide a stabilizing central mass.

\subsection{Further Exploration}

In Chapter~\ref{ch:paper_iii}, bar formation is not the focus of the study. Now that we have a sample of bars forming in different cosmological environments, we can position the disks in the parameter space described in Chapter~\ref{ch:paper_ii}. A proper synthesis with previous literature on this topic should be performed.

An interesting feature of our simulations is the existence of ``banana bars" in some of the simulations. Understanding why we get banana bars in some simulations versus a more traditional X-shaped pattern might provide an observational constraint for the Milky Way. This topic should be explored in more detail.

It would also be interesting to see how implementing a hydrodynamical, growing disk described in \S\ref{ssec:conclusions_insertion} would change our results. An interesting conclusion from \citet{aumer_2016} was that bar formation could be suppressed by GMCs. It is worth noting that these simulations also included a compact bulge, and the result might be different for a pure disk-halo system in a fully cosmological environment. The author's view is that moving in this direction is the next logical step for disk-insertion schemes.

\section{Vertical Structure}
\subsection{Summary}

The main focus of Chapter~\ref{ch:paper_iii} is how disks form vertical structure in $\Lambda$CDM environments. We showed that there are a wide variety of Milky Way-mass halos that form vertical structure consistent with structure observed in the solar neighborhood. In particular, we showed that this structure can be formed without a massive satellite encounter in a cosmological setting. This position is bolstered by the fact that we see Monoceros-like structure in all twelve of our simulations.

We also find that consistent perpendicular torques on the disk are required to explain features in the outer disk. It is unlikely that substructures far below the $10^{11}$ solar mass regime are responsible for stars stripped to higher latitudes. Global tides from disk-halo axis misalignment seem to be the most effective way to create these populations of stars. We therefore suggest that streams originating from the outer disk may primarily depend on the Milky Way's interaction with the LMC. A massive merger like the one with the LMC can cause consistent perpendicular torques its own tides or by disturbing the existing smooth Milky Way halo mass.

\subsection{More Accurate Milky Way Comparisons}

A key flaw in this work is that we had no LMC analogues. An interesting experiment would be to compare the vertical structure caused by a future LMC merger and the vertical structure generated when a disk is misaligned in a triaxial halo. On the whole, future work should proceed in the same way as the work in Chapter~\ref{ch:paper_ii}. We should seek to relate the halo environments to a common parameter space.

@@

%This thesis considered a very specific approach to understanding a very specific concept: the collisionless evolution of systems with stellar disks and dark matter halos. It is tremendously humbling to note that we never even considered gas dynamics, a cosmology other than $\Lambda$CDM, or even systems with a strong, pre-existing bulge. In the end, we were able to make some elucidating comments on the nature of stellar dynamics in $\Lambda$CDM cosmology, but no more.

This work presents an improved disk insertion scheme with applications to Galactic astronomy. We have successfully shown that we can set up and understand disks inserted into cosmological halos. Future work is suggested, and we have no shortage of potential projects to improve our scheme. The long-term goal of this work is to bridge the gap with \textit{ab initio} simulations, and we have presented a clear roadmap for doing so.

As simulation resolutions and computing power improve, we may be forced to revisit the fundamentals at the beginning of Chapter~\ref{ch:background}. Simulation codes may need to be redesigned to accommodate ever-increasing needs to relate them to observations. Our hope is that disk insertion schemes will be at the heart of these discussions as a valuable tool in understanding observational data.

Our work represents a tangible leap forward, however small, in the field of Galactic dynamics.  We hope that our work continues to inspire new developments, new comparisons, and more questions. In particular, we hope to see our work mature, and possibly tell us about the nature of the dark Universe in which we live. This is an incredibly exciting time for the field, one where we are proud to have played a small role.

### 2019-07-19 Thesis/Introduction.tex
Night before submission. Likely final pre-submission draft. Checked for spelling issues.

A model of the Universe one might consider is one comprised of cold dark matter, that is dark matter which is massive and slow, and one which has dark energy, described on large scales by a single parameter, $\Lambda$. This paradigm is called $\Lambda$CDM and is the prevailing model in the astronomy community. The model has been enormously successful in explaining observations, including the CMB power spectrum. While there remain open problems with $\Lambda$CDM, it appears to predict many properties on Galactic and extragalactic scales correctly \citep{kolb_turner,dodelson, BT}. Proposals to modify $\Lambda$CDM tend to present only slight variations on the model, such as allowing for self interacting dark matter (SIDM).
\begin{figure}
	\includegraphics[width=\textwidth]{../figures/local_group}
	\caption{A map of the Local Group obtained per the license in \citet{local_group_map}.}\label{fig:local_group}
\end{figure}

\subsection{The Story of the Milky Way}
Taking $\Lambda$CDM as the ground truth, we can begin to understand how galaxies evolved from tiny villages. Dark matter, which does not interact through any force but gravity, retains a perturbed structure after inflation. The density perturbations collapse, forming potential wells of dark matter \citep{taylor_2011}. The baryons, interacting through baryon acoustic oscillations (BOA), are delayed in falling into the potential wells made by the dark matter. As the baryons fall into the centers of the potential wells over time, the cores of the dark matter distributions, comprised of spheroidal units called halos, contract. After infall into a dark matter halo, the baryons cool to the point where stars are able to form. This is how the first galaxies formed.

@@

Simulations are the complementary theoretical understanding of galactic and extragalactic systems. When a simulation is performed, theoretical input is considered. This includes a model of cosmology, or a model of the galaxy, the nature of dark matter, etc. All of this knowledge is reduced to a fluid system which may be solved through numerical techniques on a computer.

@@

When the results of a simulation are studied, we are studying the output of a given set of assumptions. In principle, this should allow us to evaluate whether or not cosmology is consistent with dynamics on a Galactic scale. $\Lambda$CDM has been tested in this way for decades, and we detail some global results in what follows. In particular, we focus on where simulations have shown discrepancies in cosmological theory.

We broadly commented on the hierarchical formation of galaxies. Over time, galaxies that merge with other galaxies should become well-mixed in the host halo. That is, there should be an underlying smooth distribution of mass in which all of the subhalos reside. A key prediction of $\Lambda$CDM is that this smooth distribution should have a universal density profile in the limit that dark matter dominates the Universe \citep{nfw}. Furthermore, the dark matter should have a density distribution somewhat resembling a squashed football, a spheroid flattened on two axes. Whether or not this picture holds observationally in real galaxies is an open problem. The density profile proposed by \citet{nfw} is cuspy at the center, whereas some observations are more consistent with a flatter central density \citep{de_blok_core_cusp}. Astronomers have dubbed this the Core-Cusp Problem. From simply running a dark matter only simulation, we can infer the relative significance that baryonic physics plays in the central part of galaxies from these observational inconsistencies.

@@

In the hierarchical formation of galaxies, smaller galaxies merge with larger galaxies and become a part of the larger galaxy's halo. As subhalos become more a part of the host system, they experience tidal forces that stretch them apart. Dark matter and stellar material is stripped from these merging galaxies over time to form streams of tidal debris. In the long run, this debris will equlibrate with the stellar halo. It should be noted that the same effects apply to globular clusters too.

At the beginning of a merger event, the stellar debris left behind can detected. This is no easy task, and the first stream for which this accomplished was the debris of the Sagittarius Dwarf Spheroidal galaxy (Sgr dSph). The Sgr dSph was discovered by \citet{ibata_discovery}. The corresponding stream was later detected by \citet{newberg_2002} and \citet{majewski_2003}.

@@

More streams have been discovered since then, and a few notable ones are the Orphan Stream \citep{grillmair_2006,belokurov_2007,newberg_2010}, remarkable for its apparent lack of progenitor, TriAnd, A13, and the Monoceros Ring. This is by no means a complete accounting of known stellar streams. Something on the order of more than a dozen streams are known to exist in the Milky Way alone \citep{sanders_binney_2013_a}. While streams alone have been useful in getting some broad constraints on the Galactic potential and geometry of the halo, complications arise from the fact that streams do not follow orbits exactly \citet{sanders_binney_2013_a} and the fact that the Milky Way potential is predicted from theory to be time dependent.

@@

The general idea for the modification is to divide a dark-matter-only (DMO) zoom-in simulation into three phases.  During the first, a cosmological simulation of a region of interest is run, and a halo is identified for study.  A rigid disc potential is then slowly grown in this halo, allowing the halo particles to respond adiabatically to its time-varying potential.  Finally, the rigid disc is replaced by live particles and the final phase of the simulation proceeds with live disc and halo particles.  \citet{debuhr_2012} use this scheme to introduce stellar discs into dark matter halos from the Aquarius Project \citep{springel2008}.  They add a rigid disc potential at redshift $z=1.3$ with a mass parameter for the disc that grows linearly with the scale factor until $z=1$ when it reaches its final value.  The disc potential is initially centered on the potential minimum of the halo and oriented so that its symmetry axis points along either the minor or major axis of the halo.  During the rigid disc phase, the centre of the potential moves due to its interaction with the halo (Newton's 3rd law) so that linear momentum is conserved.  However, the orientation of the disc is held fixed and therefore angular momentum isn't.  To initialize the live disc, \citet{debuhr_2012} approximate the halo potential as a flattened logarithmic potential and then determine the disc distribution function by solving the Jeans equations in axisymmetry.

@@

\citet{bauer2018a} improved on the scheme even further by increasing the realism of the growth phase by modelling the disc as a rigid body. The results they found were qualitatively similar to \citet{debuhr_2012} and \citet{ys_2015}  in terms of disc dynamics. The live phase was also improved by using the \textsc{GalactICS} code which actually samples a three-integral DF. It is using this tool that this thesis is constructed upon.

\section{Organization of Thesis}

The thesis is organized in the following manner. In Chapter~\ref{ch:background}, we reduce the necessary theoretical background to understand this work to a succinct summary. Following this, in Chapter~\ref{ch:paper_i}, we explore the first peer-reviewed paper on which this thesis is based. It outlines the disk insertion scheme and comments on how the dynamic nature of the Milky Way's stellar disk should affect its inner halo. Chapter~\ref{ch:implementation} is intended to clarify some implementation details of the algorithm in Chapter~\ref{ch:paper_i}. Chapter~\ref{ch:paper_ii} presents the second published paper which focuses on the question of how stable disks are against forming bars in a cosmological setting. A final paper is presented in Chapter~\ref{ch:paper_iii}, where we look at how disks form vertical structure in a cosmological setting. We conclude with summaries and ideas for future work in Chapter~\ref{ch:conclusion}.

>>>>

To understand how dark matter is interacting with galaxies, we need to pin down its dynamics. This depends on what dark matter actually is. Some candidates, such as neutrinos, are relativistic particles that constitute hot dark matter. Warm and cold dark matter candidates become progressively non-relativistic, and there are many candidate particles for these models. An interesting and currently active proposal is that massive WIMPs, which were relativistic and coupled to baryons at early times, but cooled as the Universe expanded.

A model of the Universe one might consider is one comprised of cold dark matter and dark energy, described on large scales by a single parameter, $\Lambda$. This paradigm is called $\Lambda$CDM and is the prevailing model in the astronomy community. The model has been enormously successful in explaining observations, including the CMB power spectrum. While there remain open problems with $\Lambda$CDM, it appears to predict many properties on Galactic and extragalactic scales correctly \citep{kolb_turner,dodelson, BT}. Proposals to modify $\Lambda$CDM tend to present only slight variations on the model, such as allowing for self interacting dark matter (SIDM).
\begin{figure}
	\includegraphics[width=\textwidth]{../figures/local_group}
	\caption{A map of the Local Group obtained per the license in \citet{local_group_map}.}\label{fig:local_group}
\end{figure}

%\subsection{The Story of the Milky Way}
Taking $\Lambda$CDM as the ground truth, we can begin to understand how galaxies evolved from tiny villages. Dark matter, which does not interact through any force but gravity, retains a perturbed structure after inflation. The density perturbations collapse, forming potential wells of dark matter \citep{taylor_2011}. The exact manner in which this happens is \textit{linear}, meaning that the amplitude of the perturbations increases proportionally with the expansion of the Universe, as measured by the scale factor.

The baryons, interacting through baryon acoustic oscillations (BOA), are delayed in falling into the potential wells made by the dark matter. As the baryons fall into the centers of the potential wells over time, the cores of the dark matter distributions, comprised of spheroidal structures called halos, contract. After infall into a dark matter halo, the baryons cool to the point where stars are able to form. This is how the first galaxies formed.

@@

Simulations link theory and observations, allowing the theory to be tested. When a simulation is performed, theoretical input is considered. This includes a model of cosmology, or a model of the galaxy, the nature of dark matter, etc. All of this knowledge is reduced to a fluid system which may be solved through numerical techniques on a computer.  The output of the simulation can then be compared to observations to test the underlying theory's validity.

@@

This thesis studies dynamics on a Galactic scale. Through simulations of galaxies, we can understand how dark matter influences their evolution. $\Lambda$CDM has been tested in this way for decades, and we detail some global results in what follows. In particular, we focus on where simulations of galaxies have shown discrepancies in cosmological theory.

We broadly commented on the hierarchical formation of galaxies. Over time, galaxies that merge with other galaxies should become well-mixed in the host halo. That is, there should be an underlying smooth distribution of mass in which all of the subhalos reside. A key prediction of $\Lambda$CDM is that this smooth distribution should have a universal density profile in the limit that dark matter dominates the Universe \citep{nfw}. Furthermore, the dark matter should have a density distribution somewhat resembling a squashed football, a spheroid flattened on two axes. Whether or not this picture holds observationally in real galaxies is an open problem. The density profile proposed by \citet{nfw} is cuspy at the center, whereas some observations are more consistent with a flatter central density \citep{de_blok_core_cusp}. Astronomers have dubbed this the Core-Cusp Problem.

@@

In the hierarchical formation of galaxies, smaller galaxies merge with larger galaxies and become a part of the larger galaxy's halo. As subhalos become more a part of the host system, they experience tidal forces that stretch and squeeze their mass content. Dark matter and stellar material is stripped from these merging galaxies over time to form streams of tidal debris. In the long run, this debris will equlibrate with the stellar halo. It should be noted that the same effects apply to globular clusters too.

Baryonic debris left behind can detected. The first stream for which this accomplished was the debris of the Sagittarius Dwarf Spheroidal galaxy (Sgr dSph), a dSph galaxy discovered by \citet{ibata_discovery}. The corresponding stream was later detected by \citet{newberg_2002} and \citet{majewski_2003}.

@@

More streams have been discovered since then, and a few notable ones are the Orphan Stream \citep{grillmair_2006,belokurov_2007,newberg_2010}, remarkable for its apparent lack of progenitor, TriAnd, A13, and the Monoceros Ring. This is by no means a complete accounting of known stellar streams. Something on the order of nearly a hundred known to exist in the Milky Way alone \citep{sanders_binney_2013_a, ibata_2019}. While streams alone have been useful in getting some broad constraints on the Galactic potential and geometry of the halo, complications arise from the fact that streams do not follow orbits exactly \citet{sanders_binney_2013_a} and the fact that the Milky Way potential is predicted from theory to be time dependent.

@@

The general idea for growing a disc is to divide a zoom-in simulation into three phases.  During the first, a cosmological simulation of a region of interest is run, and a halo is identified for study.  A rigid disc potential is then slowly grown in this halo, allowing the halo particles to respond adiabatically to its time-varying potential.  Finally, the rigid disc is replaced by live particles and the final phase of the simulation proceeds with live disc and halo particles.  \citet{debuhr_2012} use this scheme to introduce stellar discs into dark matter halos from the Aquarius Project \citep{springel2008}.  They add a rigid disc potential at redshift $z=1.3$ with a mass parameter for the disc that grows linearly with the scale factor until $z=1$ when it reaches its final value.  The disc potential is initially centered on the potential minimum of the halo and oriented so that its symmetry axis points along either the minor or major axis of the halo.  During the rigid disc phase, the centre of the potential moves due to its interaction with the halo (Newton's 3rd law) so that linear momentum is conserved.  However, the orientation of the disc is held fixed and therefore angular momentum isn't.  To initialize the live disc, \citet{debuhr_2012} approximate the halo potential as a flattened logarithmic potential and then determine the disc distribution function by solving the Jeans equations in axisymmetry.

@@

\citet{bauer2018a} improved on the scheme even further by increasing the realism of the growth phase by modelling the disc as a rigid body. The results they found were qualitatively similar to \citet{debuhr_2012} and \citet{ys_2015}  in terms of disc dynamics. The live phase was also improved by using the \textsc{GalactICS} code which actually samples a three-integral DF. This thesis is largely constructed based on this tool.

\section{Organization and Findings of this Thesis}

By applying our disc insertion tool, we address three key scientific questions. The first is how growing a disc in a cosmological halo affects the halo itself. Here, we attempt to understand how the way we grow the disc affects halo properties. The second is how bars form in cosmological environments. Previous studies using disc insertion techniques found that stellar discs all form very large bars. Large central bulges were found in \citet{ys_2015} to have an ameliorating affect. We explore the reasons why their disc-halo systems formed large bars, and show how we can form bars similar to the Milky Way's without including a large central bulge. The last topic we consider is how vertical structure is excited in the Milky Way. We explore the combined effect of mechanisms in the cosmological environment, and comment on how significant factors like substructure, tides, and bar buckling are for the Milky Way.

The thesis is organized in the following manner. In Chapter~\ref{ch:background}, we reduce the necessary theoretical background to understand this work to a succinct summary. Following this, in Chapter~\ref{ch:paper_i}, we explore the first peer-reviewed paper on which this thesis is based. It outlines the disk insertion scheme and comments on how the dynamic nature of the Milky Way's stellar disk should affect its inner halo. Chapter~\ref{ch:paper_ii} presents the second published paper which focuses on the question of how stable disks are against forming bars in a cosmological setting. A final paper is presented in Chapter~\ref{ch:paper_iii}, where we look at how disks form vertical structure in a cosmological setting. We conclude with summaries and ideas for future work in Chapter~\ref{ch:conclusion}.
